When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-03-19 doc/paper/vpd_et_paper.tex
finished new introduction

The opposing effects of increased atmospheric demand and higher stomatal stress lead to two possible perspectives for how evapotranspiration (ET) responds to shifts in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand and fails to account for ecosystem response \citep{Swann_2016}. In reality, plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. This leads to a plant physiology perspective, in which an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure.  In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET response.

@@

The objective of the present manuscript is to evaluate the VPD dependence of ET, in non-extreme soil drought conditions. The goal of this paper is to use reasonable approximations as a tool to develop intuition for plant response to atmospheric drying. This intuition will aid interpretation of observations and full complexity models. In order to quantify plant response to perturbations in VPD, we apply a Penman-Monteith framework to derive the theoretical response of ET to VPD. The model is then validated and tested at multiple eddy-covariance stations spanning various climates and plant functional types.

>>>>

The opposing effects of increased atmospheric demand and higher stomatal stress lead to two possible perspectives for how evapotranspiration (ET) responds to shifts in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand and fails to account for ecosystem response \citep{Swann_2016}. In reality, plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness \citep{Farquhar_1978, Ball_1987, Leuning_1990, MEDLYN_2011}. This leads to a plant physiology perspective, in which an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure.  In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET response.

@@

The objective of the present manuscript is to evaluate the VPD dependence of ET, in non-extreme soil drought conditions. The goal of this paper is to use reasonable approximations as a tool to develop intuition for plant response to atmospheric drying. This intuition will aid interpretation of observations and full complexity models. In the past, similiar approaches were used to understand interactions between stomatal conductance and evapotranspiration \citep[e.g.,][]{Jarvis_1986, Mcnaughton_1991}. However, at the time researchers' understandng of the form of VPD's effect on plant physiology was limited, so they could not explore the sensitivity of ET to VPD, including VPD's effect on stomatal conductance and plant function.

Recent results have drastically improved our understanding of VPD's impact on physiology. \citet{MEDLYN_2011} developed a model for leaf-scale stomatal conductance ($g_s$), including VPD response, by combining an optimal photosynthesis theory \citep{Farquhar_1980, Katul_2010} with an empirical approach, and extended this model to the ecosystem scale in \citet{Medlyn_2017}. Additionally, \citet{Zhou_2014} demonstrated that a new quantity $uWUE$ ($\frac{GPP\; \sqrt{vpd}}{ET}$) properly captures a constant relationship between GPP, ET, and VPD over a diurnal cycle. uWUE is also remarkably well conserved in the growing season across space and time, within a PFT \citep{Zhou_2015}. These results provide novell tools to examine how ET varies in response to VPD, while explicitly including VPD's effect on stoamtal conductance. In this manuscript, we derive the theoretical response of ET to VPD using a Penman Monteith framework. Our theory is validated and tested at multiple eddy-covariance stations spanning various climates and plant functional types.
